\SW: II. New lens candidates from the CFHTLS discovered through citizen science

### 2015-03-17 doc/sw-cfhtls.tex

A good training sample consists of a representative set of objects that
one wants to find and another set of objects which appear to be from the
former set but are of different origin in reality and which one can
learn to discard efficiently. Indeed, we wanted to have a good training
sample for the \sw users so that they can correctly identify the true
lens candidates. Hence, in addition to the simulated lenses, we
added a sample of duds and false positives to the training sample. Duds
are images which have been visually inspected by experts and confirmed
to contain no lenses.  False positives are systems which look like
lenses but are not, for example, spiral galaxies, starforming galaxies,
chance alignments of features arranged in a lensing configuration and
stars.

@@

We find that about 40\% of the known confirmed sample of lenses are
missed (see \Tref{tab:stats}). While this seems like a fraction, we note
that many of the missed lenses are from the \rf sample with small
Einstein radii and faint lensed images (see \Fref{fig:compre}). Among
the confirmed lenses from the \rf, about 50\% are missed. Out of the
missed sample of 18 lenses, about half of them are visually difficult to
detect and the other half appear to have faint blue smudges around
galaxies which should have been easier to identify. Similarly,
if we consider the \af lens sample, $\sim$23\% are missed by \sw. This
is a relatively small sample of $\sim 5$ systems and visual inspection
suggests that, by and large, either the lensed features are faint
or they have odd properties which makes them difficult to identify
correctly.  ({\bf show some
example images XXX}). For further tests, we combine the \rf and \af sample.

@@

First, we check if the Nclass is different for the detected and the
missed sample. Surprisingly, most of the lenses in the known sample have
few classifications (Nclass<10) which includes both the detected and
missed lenses. Many of the remaining have between 10 to 20
classifications. And, only a few subjects have Nclass>20 which continue
to remain for long in the database because these candidates may have
been difficult to identify.

@@

From the SWAP, the image cutouts inspected by the \sw volunteers receive
a status of detected (if P$>$P$_{\rm acc thresh}$), rejected (if
P$<$P$_{\rm rej thresh}$) and undecided (if P$_{\rm rej thresh}$ $<$ P
$<$ P$_{\rm acc thresh}$). In \Fref{fig:comppos}, we compare the
positions of simulated lenses which are detected (red), undecided
(green) and rejected (blue). We also make the same comparison for the
known lens candidates in the right panel.  We note that for some cases
randomly selected subsamples are shown for the ease of visual
comparison. For example, the absolute number of simulated lenses
detected is too high compared to the rejected sample. We do not find any
strong visual correlation in the rate of detections as a function of the
position for both the simulated and the known lens sample. Thus, the
completeness of the lens sample does not seem to be significantly
affected by whether a lens is located close to the border or well within
the center.

@@

Here, we investigate if the skill or experience of volunteers is
systematically different between the sample of detected and missed
lenses.

We check how the posterior probability P (see XXXX Paper I) of an image
or a subject to contain a lens changes as the image receives more
classifications from multiple users.  In \Fref{fig:detmis}, we show the
trajectory plots of a few examples of detected lenses (top left panels)
and missed lenses (bottom left panels) by \sw at stage I. The number of
classifications (Nclass) for a subject increase from top to bottom.
Also, every subject is assigned a prior probability P0=2.e-4 (grey
dashed line) and starts at the middle of the trajectory plot. The P
value of a subject is updated with every classification from the
volunteer.  If a volunteer identifies a lens candidate, the trajectory
moves to the right otherwise moves to the left. A subject is accepted if it
crosses the blue-dashed line marking the (P$_{\rm acc thresh}=0.95$) on
the right and is rejected, if it crosses the red-dashed line marking the
P$_{\rm rej thresh}=1.e-7$ on the left.

By how much the P value will change depends on how well the
volunteers are performing on the training sample. Thus, more skilled
volunteers will change the P by a large factor compared to the less
skilled volunteers. This is evident in the trajectory plots as large and
small distances in between consequent points.

@@

\begin{figure*}
\begin{center}
\includegraphics[scale=0.6]{sw-cfhtls-figs/det_mis_traj.pdf}
\includegraphics[scale=0.6]{sw-cfhtls-figs/det_mis_hist.pdf}
\caption{ \label{fig:detmis}
Example cases of trajectories of known lenses on the left and histogram
of the known lenses. The trajectories on the left are lenses that are
detected by \sw are at the top and lenses that are missed are at the bottom. }
\end{center}
\end{figure*}

@@

\section{Summary and Conclusions}
\label{sec:conclude}
In this paper, we describe the framework and procedure used to generate
simulated lens sample for the blind lens search in the \cfhtls survey in
collaboration with \sw. The aim of this lens search is to find lenses that have
been missed by lens finding algorithms. The simulated lens sample is used for
training the citizen scientists, calibrating their performance and rejecting
unlikely lenses from the sample based on the classifications of citizen
scientists. As a result, the simulated lenses need to consist of realistic
looking lenses.

We use the photometric and redshift catalogues for the foreground galaxies and
additionally, color catalogues for background galaxies and quasars. We
use scaling relations to relate light properties to properties such as mass and
size to generate lens models. We further account for the
instrumental effects such as seeing and noise before creating the final lensed
images of model background sources. We add the lensed images on top of the real
galaxies and groups in the \cfhtls data in all filters.

@@

\item Crowd-sourced gravitational lens detection works, as shown in by comparing with real lenses in CFHTLS:
Real (robotically-detected and expert-confirmed) lenses are
recovered/missed at similar/comparable/different rates C’\% and P’\% - this is a partial test of supervised vs unsupervised learning

%Which real lenses were missed? False negatives

\item We found a sample of new gravitational lens candidates. An
expert-graded sample of 74 with 44 promising and 30 low probability
candiates.

The SW lenses are different from the robotically (\rf and \af) detected lenses, in the following ways.
XXXXX

@@

>>>>

A good training sample consists of a representative set of objects that
one wants to find and another set of objects which appear to be from the
former set but are of different origin in reality and which one can
learn to discard efficiently. Indeed, we wanted to have a good training
sample for the \sw users so that they can correctly identify the true
lens candidates. Hence, in addition to the simulated lenses, we
added a sample of duds and false positives to the training sample. Duds
are images which have been visually inspected by experts and confirmed
to contain no lenses.  False positives (FPs) are systems which look like
lenses but are not, for example, spiral galaxies, starforming galaxies,
chance alignments of features arranged in a lensing configuration and
stars.

@@

We find that about 40\% of the known confirmed sample of lenses are
missed (see \Tref{tab:stats}). While this seems like a fraction, we note
that many of the missed lenses are from the \rf sample with small
Einstein radii and faint lensed images (see \Fref{fig:compre}). Among
the confirmed lenses from the \rf, about 50\% are missed. Out of the
missed sample of 18 lenses, about half of them are visually difficult to
detect and the other half appear to have faint blue smudges around
galaxies which should have been easier to identify. Similarly,
if we consider the \af lens sample, $\sim$23\% are missed by \sw. This
is a relatively small sample of $\sim 5$ systems and visual inspection
suggests that, by and large, either the lensed features are faint
or they have odd properties which makes them difficult to identify
correctly. For further tests, we combine the \rf and \af sample.

@@

First, we check if the Nclass is different for the detected and the
missed sample. Surprisingly, most of the lenses in the known sample have
few classifications (Nclass$<$10) which includes both the detected and
missed lenses. Many of the remaining have between 10 to 20
classifications. And, only a few subjects have Nclass$>$20 which continue
to remain for long in the database because these candidates may have
been difficult to identify.

@@

From the SWAP, the image cutouts inspected by the \sw volunteers receive
a status of detected (if P$>$P$_{\rm acc thresh}$), rejected (if
P$<$P$_{\rm rej thresh}$) and undecided (if P$_{\rm rej thresh}<$ P
$<$ P$_{\rm acc thresh}$). In \Fref{fig:comppos}, we compare the
positions of simulated lenses which are detected (red), undecided
(green) and rejected (blue). We also make the same comparison for the
known lens candidates in the right panel.  We note that for some cases
randomly selected subsamples are shown for the ease of visual
comparison. For example, the absolute number of simulated lenses
detected is too high compared to the rejected sample. We do not find any
strong visual correlation in the rate of detections as a function of the
position for both the simulated and the known lens sample. Thus, the
completeness of the lens sample does not seem to be significantly
affected by whether a lens is located close to the border or well within
the center.

@@

Here, we investigate if the skill\footnote{see the
\Aref{appendix:skill} for the choice of this term and how it is
different from that used in paper I} of volunteers is systematically
different between the sample of detected and missed lenses.

We check how the posterior probability P (see XXXX Paper I) of an image
or a subject to contain a lens changes as the image receives more
classifications from multiple users.  In \Fref{fig:detmis}, we show the
trajectory plots of a few examples of detected lenses (top left panels)
and missed lenses (bottom left panels) by \sw at stage I. The number of
classifications (Nclass) for a subject increase from top to bottom.
Also, every subject is assigned a prior probability P0$=2\times10^{-4}$ (grey
dashed line) and starts at the middle of the trajectory plot. The P
value of a subject is updated with every classification from the
volunteer.  If a volunteer identifies a lens candidate, the trajectory
moves to the right otherwise moves to the left. A subject is accepted if it
crosses the blue-dashed line marking the (P$_{\rm acc thresh}=0.95$) on
the right and is rejected, if it crosses the red-dashed line marking the
P$_{\rm rej thresh}=10^{-7}$ on the left.

By how much the P value will change depends on how well the
volunteers are performing on the training sample. Thus, more skilled
volunteers will change the P by a large factor compared to the less
skilled volunteers. This is evident in the trajectory plots as large and
small distances in between the consequent points which we refer to as
kicks. Comparison of the kick sizes between the detected
and the missed lenses suggests that the missed lenses do not have as
many volunteers giving large kicks. We also note that most of the large
kicks seen in the trajectories of the missed lenses seem to be moving
the subjects to the right. In other words, the more skilled volunteers
are mostly classifying them as subjects with lens candidates. Further
qualitative comparison and discussion of the trajectory plots is given
in \Aref{appendix:traj}.

For a quantitative comparison of the large and small kicks for the
entire samples of detected and missed known lenses, we show a plot of
histogram on the right of \Fref{fig:detmis}. Qualitatively, there are
four types of volunteers making classifications - those causing positive
large kicks (correct classifications by more skilled volunteers, those
causing positive small kicks (correct classifications by less skilled
volunteers), those causing negative small kicks (incorrect
classifications by less skilled volunteers) and those causing negative
large kicks (incorrect classifications by more skilled volunteers). The
four histograms in the figure correspond to these four types of
volunteers for each sample (that is, detected or missed). The kick size
is small if $\Delta$ log(P)$=$log (P$_{\rm current}) - $log(P$_{\rm
previous}) < \Delta$log(P)$_{\rm cut}$ (chosen to be 1.2) and is large
if greather than $\Delta$log(P)$_{\rm cut}$.

Some of the key differences are i) the ratio of positive large kicks to
positive small kicks for the detected sample is higher than the missed
sample suggesting that the fraction of more skilled volunteers is large
for the detected sample ii) in terms of Nclass, the missed sample is
dominated by volunteers causing negative small kicks whereas for the
detected sample there is a comparable contribution from three types of
volunteers - positive small and large kicks and negative small kicks.
iii) the Nclass from volunteers with negative large kicks are lower both
for the detected and the missed samples. this is consistent with our
expectation that more skilled volunteers should not be making incorrect
classifications.Therefore, lack of more skilled volunteers classifying
the missed lenses seems to be one of the major factors.

We rerun SWAP for stage I where we use classifications from volunteers
who produce $\Delta$ log(P)$>1.2$ only. This obviously means reducing
the total Nclass per subject by a large fraction. As a result, we also
need to change the P$_{\rm acc thresh}$ which is chosen to be 0.1 and we
find that about a third of the missed lenses are detected while all the
previously detected lenses remain detected too. The others simply do not
have enough classifications from volunteers producing large positive
kicks. Thus, it may be possible to detect more promising candidates if
we preferentially show the rejected systems to a few more skilled
volunteers. We understand that changing the rejection and acceptance
thresholds will likely increase the rate of FPs along with
improved completeness but most probably the increase in FPs will not be
too significant and that it will still be worth using the new strategy.

@@

\begin{figure*}
\begin{center}
\includegraphics[scale=0.6]{sw-cfhtls-figs/det_mis_traj.pdf}
\includegraphics[scale=0.6]{sw-cfhtls-figs/det_mis_hist.pdf}
\caption{ \label{fig:detmis}
Examples of trajectories of some known lenses (top-left: detected and
bottom-left: missed) at stage I and histogram of the known lenses for
different types of volunteers (right).
}
\end{center}
\end{figure*}

@@

\section{Summary and Conclusions}
\label{sec:conclude}

We report the discovery of new gravitational lens candidates from the
the first lens search through \sw. In this search, volunteers are shown
$g-r-i$ color images of random regions of the sky taken by the CFHT Legacy
Survey. The aim of this lens search is to find lenses that have been
missed by previous searches done on the \cfhtls with lens finding
algorithms. In this search, we use a training sample to train the
volunteers and calibrate their performance which helps us in more
efficient pruning of the lens candidates. The training sample has
simulated lenses, duds and FPs. In this paper, we describe the details
of the training sample generated and tuned for the \cfhtls data.
Volunteers receive feedback messages when they click or fail to click on
the training images during live classification of real images which
helps them remain focused and improve their performance over time.

The blind lens search in the \cfhtls is done in two stages: stage I and
stage II. The aim of stage I is to do fast inspection of a large sample
of $\sim$ 430 000 images and reduce them to a smaller sample of $\sim$
3000 images with interesting lens candidates. The aim of stage II is to
carefully reinspect the smaller pool of images to exclude as many FPs as
possible and reduce the sample to $\sim$~500 candidates. In the next
step, the images are graded by three lens experts (AM,AV and PJM)
leaving in systems that are mildly interesting to very promising lens
candidates.

%For the simulated sample, we use the photometric and redshift
%catalogues for the foreground galaxies and additionally, color
%catalogues for background galaxies and quasars. Scaling relations allow
%us to relate the light properties to properties such as mass and size
%to generate lens models. We further account for the instrumental
%effects such as seeing and noise before creating the simulated lensed
%features of model background sources. We add the simulated lensed
%features on top of the real galaxies and groups in the \cfhtls data in
%all filters. A sample of duds and FPs is vetted visually before
%ingesting them to the database.

@@

\item Crowd-sourced gravitational lens detection works. We present a new
sample of 61 gravitational lens candidates with average expert grade of
1.3 and above where 1-mildly interesting, 2-probably, a lens and
3-certainly, a lens.

%Real (robotically-detected and expert-confirmed) lenses are
%recovered/missed at similar/comparable/different rates C’\% and P’\% -
%this is a partial test of supervised vs unsupervised learning

\item Compared to the sample of robotically (the \rf and \af) detected
lens candidates, the \sw sample finds lens systems with similar range in
the lens redshift and the flux and the Einstein radii of the lensed
images. However, the \sw search being a purely visual one cannot find
some of the \rf-identified-lensed images with sub-arcsecond Einstein
radii because the flux of the typically faint lensed images is obscured
by the flux from bright lensing galaxies. The \rf finds these candidates
because it works with lensing galaxy-subtracted images.

\item Qualitatively, the \sw sample finds lens systems with different
types of lensing galaxies, for example, elliptical, spiral (face on and
edge on) and small red galaxies unlike those found from robotic
searches. Similarly, the lensed images too have different colors,
morphologies and sizes which are again typically missed by any given
algorithm.

\item

@@

\section{Qualitative comparison of detected and missed known lenses}
\label{appendix:traj}

\section{Skill}
\label{appendix:skill}
In paper I, the term skill is given by the expectation value of the
information gain which is related to the P$_{\rm L}$, the probability to
correctly identify a lens as a lens and P$_{\rm D}$, the probability to
correctly identify a dud as a dud. The skill does not have simple
dependency on P$_{\rm L}$ and P$_{\rm D}$. One can get the same value of
skill for different combinations of P$_{\rm L}$ and P$_{\rm D}$.

As described in the appendix of paper I, the posterior probability P is
related to the volunteers skill (or information gain). Also, a volunteer
with a given skill, will have different positive ($\Delta$~log(P)) and
negative (-$\Delta$~log(P)) kicks on the trajectory plot. Thus, it is
not useful to compare classifications of high and low skill volunteers
to understand if there is a difference in their performance and if the
high skill volunteers can increase the completeness of the lenses. For
example, a volunteer who is very good at rejecting duds and not so good
at identifying lenses will still have a high skill. But such a
volunteer will fail to detect many of the interesting lenses. Therefore,
we do not compare high skill vs. low skill volunteers to address this
issue.

For lack of a better word, we also use the same term skill which has a
similar meaning to that described in paper I but not identical. We use a
more qualitative definition implying the ability of the volunteers to
make correct lens identification.

%%%%%%%%%%%%%%
XXXXX
